---
title: "Module 2"
author: "Claire Fields"
format: html
editor: visual
---

```{r setup, include=FALSE}
# Set options, housekeeping
knitr::opts_chunk$set(
	echo = FALSE,
	include = TRUE,
	message = FALSE,
	warning = FALSE)
```
# Workshop 1
# Load the primary data science framework and Excel import library
library(tidyverse)
library(readxl)

# Practice Import A: Loading a standard comma-separated plain text file
benthic_cover <- read_csv(here::here("data/workshop1/reef_cover_log.csv"))
# :: loads a package one time. here function indicates that you are working in the project folder

# Practice Import B: Parsing a tab-separated telemetry instrument array string
acoustic_stream <- read_tsv(here::here("data/workshop1/acoustic_telemetry_stream.txt"))

# Practice Import C: Targeting a specific sheet in a multi-tab Excel spreadsheet
fisheries_annual <- read_excel(here::here("data/workshop1/fish_catch_data.xlsx"), sheet = "Commercial_2026")

# Read in mangrove_data
mangrove_data <- read_csv(file = here::here("data/workshop1/mangrove_survey_raw.csv"))
# has a bunch of notes at the top that will mess up our reading

mangrove_data <- read_csv(
  here::here("data/workshop1/mangrove_survey_raw.csv"),
  skip = 5,   # Skip the first 5 lines of field notes
  na = c(".", "NA", "9999", "ND", "blank"))  # Convert known text alts to true NA

# Tibble vs Data frame
# tibbles are automatically how your data is loaded under tidyverse
# Force a modern tibble to degrade into a legacy base R data frame structure
benthic_cover_df <- as.data.frame(benthic_cover)
# Print the old-style dataframe structure to view
print(benthic_cover_df)
# And compare with tibble alternative
print(benthic_cover)


# Wrangling with Palmer Penguins
library(palmerpenguins)
data("penguins")

# Examine the structure of the dataset - always do this when loading a new dataset!
glimpse(penguins) # tidyverse version (from dplyr package)
str(penguins) # base R version

# Generate an exploratory summary matrix
summary(penguins)

## Foundational Grammar
# Vertically slice specific morphometric variables by explicit name
morphology_metrics <- select(penguins, species, bill_length_mm, bill_depth_mm, body_mass_g)
glimpse(morphology_metrics)

# Retain a continuous block of attributes using the colon operator
spatial_block <- select(penguins, species:island)

# Discard logistics tracking attributes while preserving everything else using the minus sign
clean_scientific_fields <- select(penguins, -year)

# Isolate observations belonging to a single categorical target group
adelie_cohort <- filter(penguins, species == "Adelie")

# Sift out individuals using continuous numerical boundary thresholds
# Preserves only large penguins whose mass exceeds 4500 grams
heavy_penguins <- filter(penguins, body_mass_g > 4500)

# Combine multiple conditional parameters across separate attributes
# Preserves records matching Gentoo penguins sampled explicitly on Biscoe Island
biscoe_gentoo <- filter(penguins, species == "Gentoo" & island == "Biscoe")

# Sift records matching multiple targeting flags within an explicit set
sub_islands <- filter(penguins, island %in% c("Dream", "Torgersen"))

# Sort penguins by ascending body mass (Default setting: Smallest mass first)
lightest_first <- arrange(penguins, body_mass_g)

# Sort penguins in descending sequence using the desc() layout wrapper
heaviest_first <- arrange(penguins, desc(body_mass_g))

# Execute nested sorting criteria: Group by species, then sort by descending bill length
stratified_morphology <- arrange(penguins, species, desc(bill_length_mm))

## Pipe operator
penguins_final <- penguins |>
  mutate(bill_ratio = bill_length_mm / bill_depth_mm) |>
  filter(species == "Adelie")

## Mutate

penguin_ratios <- penguins  |> 
  mutate(body_mass_kg = body_mass_g / 1000,   # Convert grams to kilograms
         bill_ratio = bill_length_mm / bill_depth_mm  # Bill ratio
  )
glimpse(penguin_ratios)

## Data Aggregation
# Grouping our active memory penguins by species
grouped_penguins <- group_by(penguins, species)

# Notice that the table looks identical, but metadata notes 'Groups: species [3]'
print(grouped_penguins)

# Collapsing the buckets into explicit summary metrics
species_mass_summary <- summarise(grouped_penguins,
                                  mean_mass_g = mean(body_mass_g)
)

print(species_mass_summary)
# if you have missing values then mathematical aggregations will not work to protect you 
# Overcoming the missing value trap using na.rm = TRUE
biological_signal <- penguins %>%
  group_by(species, sex) %>%
  summarise(
    sample_size = n(),                                     # Count total individuals per category
    mean_mass_g = mean(body_mass_g, na.rm = TRUE),         # Calculate mean ignoring missing cells
    sd_mass_g   = sd(body_mass_g, na.rm = TRUE)            # Standard deviation calculation
  )

print(biological_signal)

# Workshop 2

Here we are going to be wrangling in messy data on the Ross River estuary fish surveys and produce final clean tables and figure. We have 4 separate files to wrangle in and use

## Housekeeping

```{r housekeeping}
#| message: true
#| warning: false
rm(list = ls())

#This clears the environment just in case something is remaining
```

## Packages

```{r}
library(tidyverse) # Inc dplyr, ggplot2, lubridate etc
library(readxl) # For reading in excel files
```

## Load files

### Main dataset

The catch data is split across 4 tabs in one excel file Load individually

```{r}
catch1 <- read_excel(here::here("data/estuary_catch_log.xlsx"), sheet = 1)
catch2 <- read_excel("../data/estuary_catch_log.xlsx", sheet = 2)
catch3 <- read_excel("../data/estuary_catch_log.xlsx", sheet = 3)
catch4 <- read_excel("../data/estuary_catch_log.xlsx", sheet = 4)
```

```{r}
#inspect these tabs
catch1
catch2
catch3
catch4
```

## Consolidate

```{r}
catch_data <- bind_rows(catch1, catch2, catch3, catch4)
# Clean up the environment
rm(catch1, catch2, catch3, catch4) # Clean up the environment
```

```{r}
catch_data
head(catch_data)
glimpse(catch_data)
str(catch_data)
catch_data|>
  distinct(species)
catch_data|>
  distinct(site)
```

##### Data Info:

-   four species but they are in common names and inconsistent
-   four sites but they are not consistently written

## Load in other data

```{r}
metadata<-read_csv("../data/estuary_metadata.csv")
dictionary<-read_csv("../data/species_dictionary.csv")
sonde_raw<-read_csv("../data/estuary_sonde_data.csv")
summary(sonde_raw)
```

```{r}
catch_data <- catch_data |>
  mutate(
    site = str_to_lower(str_trim(site)),
    site = str_replace_all(site, " ", "_")) |> 
  # Convert all chr to fct
  mutate(across(where(is.character), as_factor),
    species = str_to_lower(str_trim(species)),
    species = str_replace_all(species," ", "_")
  )

metadata <- metadata |>
  mutate(
    site_name = str_to_lower(str_trim(site_name)),
    site_name = str_replace_all(site_name, " ", "_")) |> 
  # Convert all chr to fct
  mutate(across(where(is.character), as_factor)
  )

# Taxonomic Translation
catch_data<- catch_data |>
  left_join(dictionary, by=join_by(species==common_name))
```

#### Notes:

-   changed all the naming issues
-   Environmental variables look like they are mostly in good shape (all numeric), however the turbidity data has a failed value (-999) so we'll need to deal with that

## Fix up the sonde data

```{r}
sonde<-sonde_raw|>
  mutate(
    # Parse the messy character string
    datetime=dmy_hm(timestamp),
    # Convert 999 to NA
    turbidity = na_if(turbidity, -999),
    # Convert negative salinity to NA
    salinity = ifelse(salinity < 0, NA, salinity),
    # in salinity = if salinity is less than zero make it NA if not keep it salinity
  )
# Because our catch data is on a daily basis we don't really care about having the date time split into 15 minute measurements. maybe we just get the mean with standard deviation
# Summarise 15-minute sonde data into daily means
sonde <- sonde |>
  mutate(  
    date = as_date(floor_date(datetime, "day")), # as_date changes it to the date format rather than the dttm (date time)
  ) |>
  group_by(site, date) |>
  summarise(
    mean_temp = mean(temperature, na.rm = TRUE),
    mean_salinity = mean(salinity, na.rm = TRUE),
    .groups = "drop"
  )
# join it to the catch data
catch_data<- catch_data |>
  left_join(metadata, by = join_by(site == site_name)) |>
  left_join(sonde, by=join_by(site,date))
# 
```

#### Notes

-   Salinity had negative values we deleted
-   Turbidity had NAs we removed
-   We changed from 15 minute intervals to daily average for our sonde readings, but dropped turbidity because it is not relevant to our hypothesis
-   student did not include their zero counts, we need to add them in 4 sites, 4 species, 30 days (4*4*30 = 480 observations expected)

## Implicit zeros

```{r}
catch_data<-catch_data |>
  complete(
    nesting(site, date, lat, lon, zone, mean_temp, mean_salinity), scientific_name) |>
  mutate(
    count=coalesce(count,0))
```

Data wrangling complete!

## Phase 4: Statistical Exploration

```{r}
# Are there differences in environmental variables among zones? Otherwise, what is the point in testing their effect?

# Salinity
salinity_summary <-
  catch_data|>
  group_by(zone) |>
  summarise(
    sem_salinity = sd(mean_salinity) / sqrt(n()),
    sd_salinity = sd(mean_salinity, na.rm = TRUE),
    median_salinity = median(mean_salinity, na.rm = TRUE),
    .groups = "drop"
  )

#plot
salinity_summary |> 
  ggplot(aes(x = zone, y = median_salinity)) +
  geom_point() +
  geom_errorbar(aes(ymin = median_salinity - sd_salinity, ymax = median_salinity + sd_salinity), width = 0.2) +
  theme_minimal() +
  labs(#title = "Median Salinity by Zone with Standard Deviation Error Bars",
       x = "Zone",
       y = "Median Salinity (ppt)")

# Temperature
temperature_summary <- 
  catch_data |>
  group_by(zone) |>
  summarise(
    sd_temperature = sd(mean_temp, na.rm = TRUE),
    median_temperature = median(mean_temp, na.rm = TRUE),
    .groups = "drop"
  )

temperature_summary |>
  ggplot(aes(x = zone, y = median_temperature)) +
  geom_point() +
  geom_errorbar(aes(ymin = median_temperature - sd_temperature, ymax = median_temperature + sd_temperature), width = 0.2) +
  theme_minimal() +
  labs(#title = "Median Temperature by Zone with Standard Deviation Error Bars",
       x = "Zone",
       y = "Median Temperature (°C)")
```

#### Notes

Not a huge difference in temperature, but salinity has clear differences among zones.

```{r}
# Nice publication friendly plot

#| fig-cap: "Salinity Gradient across zones. Ross River Estuary System, Townsville, NQ, Australia May 2026."

salinity_summary |> 
  ggplot(aes(x = zone, y = median_salinity)) +
  geom_point() +
  geom_errorbar(aes(ymin = median_salinity - sd_salinity, ymax = median_salinity + sd_salinity), width = 0.2) +
  theme_minimal() +
  labs(#title = "Median Salinity by Zone with Standard Deviation Error Bars",
       x = "Zone",
       y = "Median Salinity (ppt)")

# abundance patterns
fish_stats <- catch_data |>
  group_by(zone, scientific_name) |>
  summarise(
    mean_catch = mean(count),
    sem_catch = sd(count) / sqrt(n()),
    .groups = "drop"
  )
```

## Phase 5: Visual Communication

```{r}
fish_stats |> 
  ggplot(aes(x = zone, y = mean_catch, fill = scientific_name)) +
  geom_line(group = 1) +
  # add in points from master data
  geom_point(data = catch_data, aes(x = zone, y = count),
             position = position_jitter(width = 0.2), alpha = 0.2, size = 1, inherit.aes = FALSE) + 
  geom_point(position = "dodge", size = 4, pch = 23) +
  # SE pretty narrow, potentially not worth plotting but let's viewer see that included
  geom_errorbar(aes(ymin = mean_catch - sem_catch, ymax = mean_catch + sem_catch), 
                position = position_dodge(0.9), width = 0.2) + 
  facet_wrap(~ scientific_name, scales = "fixed") +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "italic", size = 10, colour = "grey20"), # spp names in italic
   # legend.position = "bottom"
    legend.position = "none" # legend unnecessary
  ) +
  labs(
    #title = "Predatory Fish Abundance across Estuarine Salinity Gradients", # default ggplot title appears on top
    #subtitle = "Ross River Estuary System, May 2026",
    x = "Spatial Zone",
    # scientific label for y axis of n day -1 use regex to get superscript
    y = expression("Abundance (n day"^-1*")"),
  #  y = "Daily abundance (n day^-1)"
    y = "Abundance (per day)"
  )
```
